A state space model for characterizing the trajectory dynamics of replay

### 2020-04-24 Article.tex
Update on Overleaf.

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
During sleep and immobility, hippocampal place cells fire in sequences consistent with temporally compressed versions of trajectories previously run by the animal. These replayed sequences are hypothesized to be an important mechanism for the retrieval of spatial memory in service of consolidation and decision-making. Replay events are typically evaluated based on whether they activate sequences of place cells that represent spatially continuous trajectories through the environment, but recent work has shown that these events can have more complex dynamics. For example, sequences can alternate between hovering on a particular spatial location and continuous movement or can represent continuous trajectories in other spatial environments, which may appear spatially incoherent in the context of the current environment.

To quantify the structure of replay events, we develop a state space model that uses a combination of discrete and continuous latent state to decompose place cell sequences into categories based on their latent dynamics. Each discrete latent “category” is associated with a type of continuous latent dynamic—hovering in place, spatially fragmented or spatially continuous. This allows for (1) direct comparison between different categories of sequence dynamics, (2) expression of our confidence in one or more categories explaining the data, and (3) characterization of the transitions between categories. In addition, the model can function in 2D, avoiding linearization errors on more complicated environments. We demonstrate the utility of this model on simulated and real data of an animal performing a spatial memory task.

@@

\section*{Introduction}
The hippocampus is a brain area that is important for short term learning and memory. Neurons in the hippocampus preferentially fire in response to specific locations in an environment, which results in sequences of cells firing as an animal moves through an environment that reflect the trajectory of the animal. When the animal is asleep or immobile, cells also fire in sequence, forming trajectory-like representations that can recapitulate sequences previously observed when the animal was running. These sequences are thought to be important for memory consolidation or prospective decision making. (cite some examples). Therefore, characterizing the dynamics of hippocampal sequences is important for understanding memory consolidation and prospective decision making.

Current methods analyze hippocampal sequences by assuming a constant velocity trajectory \cite{DavidsonHippocampalReplayExtended2009}, corresponding to a single dynamic. This is done by computing the probability of position at each time using the place field information from the population of neurons and then fitting a line to maximize the probability of position over time. If the probabilities near the best fit line are high relative to shuffled distribution, the replay is analyzed. Often, trajectories that travel less than a certain distance \cite{PfeifferAutoassociativedynamicsgeneration2015} are excluded.

However, there is a growing evidence that trajectories that have more complex dynamics. Farooq and Dragoi \cite{FarooqEmergencepreconfiguredplastic2019} found that juvenile rats . For example, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that trajectories alternated between stationary hovering trajectories and continuous movement trajectories on a gamma timescale.

Our mathematical model precisely specifies the dyna

\section*{Results}
\subsection*{Overview of the model}

In order to validate the model's effectiveness in detecting short time scale changes in spiking dynamics, we first simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then applied our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly. For the middle third of the sequence, the cells fire in sequential spatial order. For the last third, the cells fire in a incoherent spatial pattern.

Decoding the spiking sequence requires specifying two elements: the data model and the dynamics model. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the typical Bayesian decoder, we use small time bins (2 ms) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the 20 ms bin size used with the Bayesian decoder.

Next, we specify a dynamics model for how latent position can evolve at a given time (Figure \ref{1}C). We are interested in identifying three types of different dynamics---hover, continuous, and fragmented---which are modeled with

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v3}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the hover state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the states displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other states at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}

\subsection*{The model can find canonical continuous replays}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v3}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes (arranged top to bottom). In the top panel, 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center wells respectively. The animal's position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} 1D MAP estimate of the latent position in (A) projected back into 2D. Color indicates time. The animal's position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center well as in (A). \textbf{(D)} 1D MAP estimate of the latent position in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{Figure2}
\end{figure*}

\subsection*{The model also finds replays that would not normally be classified by the Bayesian Decoder}

\subsection*{Most replays are spatially coherent, but often not at the speed posited by canonical replay}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v3}
\caption{A-D. Four examples of replays that would not be well-characterized by using the Bayesian decoder. Figure conventions are the same as in Figure 2. \textbf{(A)} A replay that starts down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that stays at the left well for the entirety of the SWR while the animal is in the center arm. \textbf{(C)} A replay that begins stationary at the left well and then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins stationary at the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories (Hover, Hover-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Hover, Hover-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{Figure3}
\end{figure*}

\subsection*{Most replays only have one dynamic but some do not}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v3}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot of the most common combinations of dynamics within each ripple. Each row represents a possible category present in the SWR. Each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occurred as indicated by the bar plot above each category. The total number of each category is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of hovers on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of multiunit spike rates for each dynamic category within each SWR.
}
\label{Figure4}
\end{figure*}

\subsection*{Many replays have time periods where the latent position remains at just one position and that position can be both local and non-local to the animal's position}
\subsection*{Non-local hovers occur all over the track, but mostly at reward wells and choice points.}

\subsection*{Removing spatial information from the model leads to fewer classified ripples}

\section*{Discussion}

@@

\subsection*{Classification of Dynamic States}
We used a threshold of 0.8 to classify the probability of each state into 5 categories. Time periods during sharp wave ripples are labeled as Hover, Continuous, or Fragmented when the probability of each state is above 0.8. Time periods are labeled as Hover-Continuous-Mix or Fragmented-Continuous-Mix when the sum of Hover and Continuous or Fragmented and Continuous are above 0.8, respectively. Time periods where none of these criterion are met are considered unclassified because there is not strong enough evidence for one or more of the states.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1-supplemental1/Figure1_v1_supplemental1}
\caption{The model is robust to changes in the discrete transition matrix. Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value--which governs the probability of remaining in that state. The off-diagonal values--the probability of switching to one of the other states--are set to be equally likely with the remainder of the probability, as in Figure 1D. The first plot (upper left) shows the case when all states are equally likely. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero--i.e. the case where there is no probability of switching to another state.}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}

>>>>

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past or possible future experiences play a critical role in decision-making processes. The hippocampus is a brain region known to express these representations across multiple states, including during sharp-wave ripple (SWR) events where sets of hippocampal place cells can be activated in time-compressed sequences consistent with a rapid mental traversal of a path. Efforts to understand these representations typically make arbitrary assumptions about the character and dynamics of the generative representations. As a result, we lack a clear picture of the nature and diversity of SWR representations. Here we develop a flexible state space model that uses a combination of discrete and latent states to describe the content and time evolution of SWR events. Application of this model reveals that a much larger proportion of SWR events contain coherent spatial content that previously reported, and that the dynamics of events span a surprisingly wide range.

@@

\section*{Introduction}
The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These internal "generative" representations can be identified because the underlying neural activity patterns resemble those seen during actual events. In particular, an encoding model that relates neural firing to external covariates can be constructed during an experience, and this model can inverted to decode an internal representation of the covariates from the neural activity at a different time.

Mental events do not occur in the same time span as actual events; memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to those actions. Consistent with this, generative representations can unfold rapidly. As a result, decoding generative representations requires making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

Hippocampal replay events are a prototypical example of a generative representation. As animals move through space, neurons in the hippocampus preferentially fire in at specific locations in an environment, and thus sets of cells can fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous or potential future experience, and previous work has reported that these sequential firing events proceed at an average speed of ~10 meters / second, about 20x faster than the animal's usual movement speed.

Identifying these events typically involves multiple steps and assumptions about the nature of the event. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on an threshold for event size chosen by the experimenter. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed. A statistical test is then used to determining whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Moreover, there is now substantial evidence that a subset of these events express stationary representations, which are often excluded. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

 Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the weighted mean or maximum of each posterior time bin and connecting each time bin with a line (cite). For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used fixed time bins and neither took into account the uncertainty of the decoded estimates, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture different types of movement dynamics, ranging from stationary to rapidly evolving to disorganized, and provide a statistical assessment of confidence for the dynamic. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model (cite Brown, Deng) of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches (refs) and characterizes represented trajectories as a mixture of three underlying patterns of dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike wave form features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 animals, focusing on SWR events to permit a direct comparison to previous work. We find that the vast majority of SWRs contain spatially coherent content. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at speeds consistent with real animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. Our method provides a useful and powerful tool for understanding the dynamics of mental representations, and could be adapted for use in many brain areas.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v3}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the hover state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
In order to introduce and validate the model, we demonstrate how the model works on a simulated dataset (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly. For the middle third of the sequence, the cells fire in sequential spatial order. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize replay events in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model and the dynamics model. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we use small time bins (2 ms vs. 20 ms) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a dynamics model for how latent position---the "mental" position of the animal represented by the cells---can evolve in each of these sequences (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can change from the previous time step. In the stationary state, we expect latent position to not change between time steps, so we use an identity state transition matrix. In the continuous state, we expect the latent position to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since we are interested in identifying replays that move at speeds much faster than the animal's speed, we set this to 6.0 cm. This means that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or 24.5 m/s), consistent with replay speeds observed in previous studies \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}. In the fragmented state, we expect that the latent position can move to any available position. We model this using a uniform state transition matrix, which makes all positions equally likely.

Finally, we specify how likely we are to stay in a dynamic state versus changing to another state (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for the entirety of the replay, with a small probability of switching to one of the other dynamics. Therefore, we set the probability of staying in a state to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular state.

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic state. We summarize the resulting posterior with two quantities: the mixture probability of each state over time and the probability of position over time (Figure \ref{1}F, left and right plot respectively). From these two summaries, we can see that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The probability of the stationary state has high probability at the beginning of the simulated replay, reflecting the consistent spiking from one cell. This gives way to the continuous state, reflecting the trajectory-like spiking from the population of cells. Then the fragmented state dominates for the last third when the population spiking is not spatially coherent. We can also see that these three dynamics are reflected in the estimate of latent position of the animal.  Importantly, we get a much less variable estimate due to the correct specification of the dynamics.

One way to interpret the probabilities in our model is as a classifier of speeds of replay. To demonstrate this, we applied the model to simulated replays of increasing speed (Figure \ref{1}G), from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three dynamics being high probable (where we define highly probable to be over 0.8 probability), there are in-between speeds where two of the dynamics dominate with high probability; that is, the sum of two of the dynamics' probabilities is over 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary state has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its dynamics.

In order to verify that the model was robust to the choice of probability of staying in a state, we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure 1-Supplemental Figure 1). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v3}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes (arranged top to bottom). In the top panel, 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center wells respectively. The animal's position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} 1D MAP estimate of the latent position in (A) projected back into 2D. Color indicates time. The animal's position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center well as in (A). \textbf{(D)} 1D MAP estimate of the latent position in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{The model finds known continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v3}
\caption{A-D. Four examples of replays that would not be well-characterized by using the Bayesian decoder. Figure conventions are the same as in Figure 2. \textbf{(A)} A replay that starts down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that stays at the left well for the entirety of the SWR while the animal is in the center arm. \textbf{(C)} A replay that begins stationary at the left well and then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins stationary at the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories (Hover, Hover-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Hover, Hover-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{3}
\end{figure*}

Next, we wanted to validate our model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with clear sequential population activity (Figure \ref{2}A, top panel). We observed that the probability of being in the continuous state is high throughout for this SWR, with the probability of being in a stationary state being slightly higher at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory to 2D (Figure \ref{2}C) to better see the spatial trajectory. Moreover, when we apply the same model using the 2D position of the animal, we get a similar result as when we used the 1D linearized position (Figure 2-Supplemental 1A).

One problem with using spikes from cells identified using spike sorting is that there is potentially a lot of information lost by determining the identity of each cell---many spikes with position information are discarded because the spikes could not be assigned to one cell. Therefore, we wanted to use a method which directly relates multiunit spike waveform features to position without spike sorting, which we call clusterless decoding. Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016}, Kay et al. 2020). Here, if we apply it to the same SWR as the sorted spikes, we get similar decoding and classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure 2-Supplemental 1B). Note that, using clusterless decoding, the estimate of the posterior probability of position is much more narrow, indicating that there is more certainty in the true latent position at that time compared the corresponding posterior probability of position using sorted spikes (Figure \ref{2}D vs C).

\subsection*{The model finds replays with non-constant speed dynamics}
After validating our model on a single SWR, we then applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009}). We observed many replays that were classified as continuous throughout the entirety of SWR and likely would have been identified by the most commonly used Bayesian Decoder (Figure 2-Supplemental 2A-C). However, we also observed replays that did not have this structure, but could be interpreted in terms of our model. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure 3-Supplemental 1B), trajectories that stayed in one position (Figure \ref{3}B), trajectories that jumped between arms (Figure \ref{3}C-D, Figure3-Supplemental 1A, E, F), and trajectories that remained spatially incoherent throughout the SWR (Figure3-Supplemental1D). Overall, we were able to classify XX\% of ripples as containing at least one of the five dynamics. To verify that this depended on the place field coding of hippocampus, we resampled position with replacement for two epochs, breaking the relationship between spiking and spatial information. There were far fewer SWR classified when spatial information was removed (p=0.XX for epoch 1, p=0.XX for epoch 2, Figure 3-Supplemental 2).

To better understand the distribution of dynamics during SWRs, we then asked what percentage of classified replays contained spatially coherent structure---which we define as a SWR containing any of stationary, stationary-continuous mixture, or continuous categories. We found that XX\% of SWR had some spatially coherent structure while XX\% had some spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We wondered what percentage of SWR contained continuous content, like would typically be analyzed when using by the Bayesian decoder. We found that, consistent with previous reports, only XX\% had continuous content (Figure \ref{3}G). This highlights that much of the SWR replay content is missed when analyzing SWRs with the Bayesian decoder.

\subsection*{Many replay trajectories are closer to the animal's movement speed}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v3}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each ripple. Each row represents a possible category present in the SWR. Each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occurred as indicated by the bar plot above each category. The total number of each category is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of hovers on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of multiunit spike rates for each dynamic category within each SWR.
}
\label{4}
\end{figure*}

We then looked at which sets of dynamics are most common. Many, but not all, of the SWR ripples were well described by one dynamic (XX/YY, XX\%). Surprisingly, the most common SWRs were ones that were categorized exclusively as stationary-continuous mixtures (XX\%, Figure \ref{4}A). These trajectories had similar duration to the continuous events (median XX ms, Figure \ref{4}B) and occurred away from the animal (median XX cm, Figure \ref{4}C). This indicates that the trajectories of these SWRs do not represent the animal's position and move at a speed that is closer to the animal's actual movement speed when running, rather than many times the animal's movement speed.

The next most common category is those SWRs that were exclusively stationary (XX\%). Unlike the Stationary-Continuous Mixtures, most of these trajectories represented the animal's position (Figure \ref{4}C), but there were many that represented positions as far as XX cm from the animal. We call the ones that are more than 30 cm away from the animal "non-local stationary trajectories". There were XX/YY or XX\% non-local stationary trajectories. These non-local stationary trajectories represented positions all over the track, but mostly at reward wells and choice points (Figure \ref{4}D).

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v1}
\caption{Comparison of real vs. position shuffled data for two epochs from different animals. Line represents the percent of SWR classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing.}
\label{Figure5}
\end{figure}
\section*{Discussion}

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.8 to classify the probability of each state into 5 categories. Time periods during sharp wave ripples are labeled as Hover, Continuous, or Fragmented when the probability of each state is above 0.8. Time periods are labeled as Hover-Continuous-Mix or Fragmented-Continuous-Mix when the sum of Hover and Continuous or Fragmented and Continuous are above 0.8, respectively. Time periods where none of these criterion are met are considered unclassified because there is not an interpretation for all three dynamics being equally likely.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1-supplemental1/Figure1_v1_supplemental1}
\caption{The model is robust to changes in the discrete transition matrix. Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value--which governs the probability of remaining in that state. The off-diagonal values--the probability of switching to one of the other dynamics--are set to be equally likely with the remainder of the probability, as in Figure 1D. The first plot (upper left) shows the case when all dynamics are equally likely. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero--i.e. the case where there is no probability of switching to another state.}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}
